Geometric and physical interpretation of the action principle

### 2023-08-14 Articles/2022-ActionPrinciple/ActionPrincipleArchive.tex
ActionPrinciple: updating archive version

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the appendix.

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that describes how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumptions requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free,\footnote{Given that this is a three dimensional space, we can use the standard tools of vector calculus.} that is
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning\footnote{In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.}
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative\footnote{We write the exterior derivative as $\partial \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.} of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions.

@@

@@

\section*{Appendix}

In this section we will review the arguments in more detail. Given that this article is still aimed at a physics audience, the focus will be conceptual rigor, rather than mathematical rigor. What is important is to establish a precise map between physical concepts and their mathematical representation, so that it is clear what physical requirement is captured by each mathematical definition.

We divide the appendix into three sections: the first will address the single d.o.f. case, the second will give a geometric interpretation of the tools of differential geometry that is suitable for physicists, and lastly we address the multiple d.o.f. case.

@@

Now, since every functional has a corresponding form, $\partial F$ must have a corresponding $k+1$-form which must be fully identified by $\omega$. We note it as $\partial \omega$ and write
\begin{equation*}
	\partial F(\Sigma) = \int_\Sigma \partial \omega(d\Sigma) = \oint_{\sigma = \partial \Sigma} \omega(d\sigma) = F(\partial \Sigma).
\end{equation*}
This is the generalized Stokes' theorem and $\partial \omega$ is the exterior derivative of $\omega$. We use $\partial$ as we leave $d$ reserved for the infinitesimal regions.

@@

Note that we have broken down exactly how each assumption contributes to the mathematical setting. If \ref{assum_indep} fails, then $\Omega$ cannot be understood as $\omega^N$, which means we have much fewer conditions on the motion: we only know that the total flow is conserved, but we cannot break it down along each d.o.f.. On the other hand, if \ref{assum_detrev} fails we may not have a single displacement field $\vec{S}$ and/or the form that counts states would not be closed. Nonholonomic constraints will, in general, make one of the assumptions fail, for example by introducing a dependence between degrees of freedom. This type of exact mapping between assumptions and their mathematical expression is one of the key goals of Reverse Physics in particular and our overall project Assumptions of Physics more in general.

For (ii), we note that since the form is closed, we can find, at least in a contractible region, a covector potential $\theta$ such that $\omega=\partial \theta$. Those familiar with symplectic and contact geometry will know that using Darboux's theorem, we can find a set of coordinates such that we can express $\theta$ as
\begin{equation}
	\tag*{(\ref{mdof_potential_expression})}
	\theta = p_i e^{q^i} - H e^t.
\end{equation}
Just stating this is, of course, not very insightful. To gain more insight, we provide the sketch of a proof that, as we have done for the single degree of freedom, uses gauge transformations to fix the form of the potential.

As a starting point, we take conditions \ref{canonical_conditions} that state mathematically that $q^i$ and $p_i$ form independent degrees of freedom. We re-express those conditions in terms of the potential $\theta$
\begin{equation}\label{canonical_potential_conditions}
\begin{aligned}
	\omega(e^{q^i}, e^{p_j}) &= (-\partial\theta)_{q^i p_j} = -(\partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i}) = \delta^i_j \\
	\omega(e^{q^i}, e^{q^j}) &= (-\partial\theta)_{q^i q^j} = -(\partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i}) = 0 \\
	\omega(e^{p_i}, e^{p_j}) &= (-\partial\theta)_{p_i p_j} = -(\partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i}) = 0
\end{aligned}
\end{equation}
We can use our gauge freedom to set $\theta_{p_1} = 0$, much in the same way we did for the simpler case. We now have $\partial_{q^1} \theta_{p_1} = 0$ and, by the first condition, $\partial_{p_1} \theta_{q^1} = 1$. Integrating, we have $\theta_{q^1} = p_1 + g(q^i, p_2, p_3, ..., t)$ where $g$ is an arbitrary function which we can set to zero. Like in the single degree of freedom, we can do this because $g$ does not depend on $p_1$, so it can be eliminated with a gauge transformation that does not change $\theta_{p_1}$. Therefore we have:
\begin{equation}
	\theta = p_1 e^{q^1} + 0 e^{p_1} + \theta_{q^2} e^{q^2} + \theta_{p_2} e^{p_2} + ... + \theta_{t} e^{t}.
\end{equation}

@@

Comparing \ref{mdof_form_components} and \ref{mdof_Ham_eq} we see that the time components $\omega_{at}$ are actually the components of $\vec{S}$. In fact, the only non-trivial components of $\omega$ are components of $\vec{S}$. These two objects, then, are not simply related: they express the same information.

For a single degree of freedom, recalling that $S^t = 1$, we can describe $\omega$ with the following nine components
\begin{equation}\label{mdof_reduceto_sdof}
	\omega_{ab}=(-\partial\theta)_{ab} = \begin{bmatrix}
		0 & S^t & -S^p \\
		-S^t & 0 & S^q \\
		S^p & - S^q & 0
	\end{bmatrix}.
\end{equation}
In three dimensions, using the Levi-Civita symbol $\epsilon^{abc}$ we can write
\begin{equation}
	S^a = \epsilon^{abc} \omega_{bc}
\end{equation}
which tells us that $\vec{S}$ is the pseudo-vector corresponding to the direction perpendicular to $\omega$. This is the same relationship that we would have, for example, for the magnetic field psuedo-vector and the spatial components of the electromagnetic field in relativity. If we express $\omega$ in terms of $\vec{\theta}$, we have
\begin{equation}
	S^a = - \epsilon^{abc} \partial_b \theta_c
\end{equation}
which recovers \ref{sdof_displacement_potential}: the displacement field is minus the curl of $\omega$. Thus we can see exactly how the general case reduces to the single d.o.f. case.

For (iv), we simply need to apply Stokes' theorem. The situation is identical to the single degree of freedom, except that the flow now is not given by the surface integral of $\vec{S}$, but of the form $\omega$. The path $\gamma$ and its variation $\gamma'$ will form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\omega$ is closed, its surface integral depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we recover
\begin{equation}
\tag*{(\ref{mdof_action})}
\begin{aligned}
	- \int_{\Sigma} \omega(d\Sigma) &=
	\int_{\Sigma} \partial \theta (d\Sigma) \\
	&= \oint_{\partial \Sigma = \gamma \cup \gamma'} \theta(d\gamma) \\
	&= \int_{\gamma} \theta(d\gamma) - \int_{\gamma'} \theta (d\gamma') \\
	&= \delta \int_{\gamma} \theta (d\gamma).
\end{aligned}
\end{equation}

>>>>

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. [This has been the hardest problem to solve, as the modern mathematical concepts and notation have departed from physical intuition. We will therefore use, instead of the standard notation of differential geometry, an extended version of the one physicists use in general relativity.] To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will present all the key points needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the Supplementary Information.

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
	\begin{aligned}
		\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
that describes how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} In dynamical system literature, this is referred to as the vector field of the dynamical system. The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumption requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free.\footnote{Given that this is a three-dimensional space, we can use the standard tools of vector calculus.} That is, 
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning 
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. [In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.] This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero. Those familiar with differential geometry will recognize $\omega$ as the symplectic form of symplectic manifolds and of the symplectic bundles of contact manifolds.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \wedge \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions.\footnote{We write the exterior derivative as $\partial \wedge \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.}

@@

\vspace{-1mm}
\section*{Acknowledgments}
This paper forms part of the ongoing \textit{Assumptions of Physics} project \cite{aop-book}, which aims to identify a handful of physical principles from which the basic laws can be rigorously derived. The authors would like to thank Alejandro Uribe and Daniel Burns for helpful discussions.

@@

\section*{Supplementary Information}

In this Supplementary Information we will review the arguments in more detail. Given that this article is still aimed at a physics audience, the focus will be conceptual rigor, rather than mathematical rigor. What is important is to establish a precise map between physical concepts and their mathematical representation, so that it is clear what physical requirement is captured by each mathematical definition.

We divide the material into three sections: the first will address the single d.o.f. case, the second will give a geometric interpretation of the tools of differential topology that is suitable for physicists, and lastly we address the multiple d.o.f. case.

@@

Now, since every functional has a corresponding form, $\partial F$ must have a corresponding $k+1$-form which must be fully identified by $\omega$. We note it as $\partial \wedge \omega$ and write
\begin{equation*}
	\partial F(\Sigma) = \int_\Sigma \partial \wedge \omega(d\Sigma) = \oint_{\sigma = \partial \Sigma} \omega(d\sigma) = F(\partial \Sigma).
\end{equation*}
This is the generalized Stokes' theorem and $\partial \wedge \omega$ is the exterior derivative of $\omega$. We use $\partial \wedge$ as we leave $d$ reserved for the infinitesimal regions.

@@

Note that we have broken down exactly how each assumption contributes to the mathematical setting. If \ref{assum_indep} fails, then $\Omega$ cannot be understood as $\omega^N$, which means we have many fewer conditions on the motion: we only know that the total flow is conserved, but we cannot break it down along each d.o.f.. On the other hand, if \ref{assum_detrev} fails we may not have a single displacement field $\vec{S}$ and/or the form that counts states would not be closed. Nonholonomic constraints will, in general, make one of the assumptions fail, for example by introducing a dependence between degrees of freedom. This type of exact mapping between assumptions and their mathematical expression is one of the key goals of Reverse Physics in particular and our overall project Assumptions of Physics more in general.

For (ii), we note that since the form is closed, we can find, at least in a contractible region, a covector potential $\theta$ such that $\omega=\partial \wedge \theta$. Those familiar with symplectic and contact geometry will know that using Darboux's theorem, we can find a set of coordinates such that we can express $\theta$ as
\begin{equation}
	\tag*{(\ref{mdof_potential_expression})}
	\theta = p_i e^{q^i} - H e^t.
\end{equation}
Just stating this is, of course, not very insightful. To gain more insight, we provide the sketch of a proof that, as we have done for the single degree of freedom, uses gauge transformations to fix the form of the potential.

As a starting point, we take conditions \ref{canonical_conditions} that state mathematically that $q^i$ and $p_i$ form independent degrees of freedom. We re-express those conditions in terms of the potential $\theta$
\begin{equation}\label{canonical_potential_conditions}
\begin{aligned}
	\omega(e^{q^i}, e^{p_j}) &= (-\partial \wedge\theta)_{q^i p_j} = -(\partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i}) = \delta^i_j \\
	\omega(e^{q^i}, e^{q^j}) &= (-\partial \wedge\theta)_{q^i q^j} = -(\partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i}) = 0 \\
	\omega(e^{p_i}, e^{p_j}) &= (-\partial \wedge\theta)_{p_i p_j} = -(\partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i}) = 0
\end{aligned}
\end{equation}
We can use our gauge freedom to set $\theta_{p_1} = 0$, much in the same way we did for the simpler case. We now have $\partial_{q^1} \theta_{p_1} = 0$ and, by the first condition, $\partial_{p_1} \theta_{q^1} = 1$. Integrating, we have $\theta_{q^1} = p_1 + g(q^i, p_2, p_3, ..., t)$ where $g$ is an arbitrary function which we can set to zero. Like in the single degree of freedom, we can do this because $g$ does not depend on $p_1$, so it can be eliminated with a gauge transformation that does not change $\theta_{p_1}$. Therefore we have:
\begin{equation}
	\theta = p_1 e^{q^1} + 0 e^{p_1} + \theta_{q^2} e^{q^2} + \theta_{p_2} e^{p_2} + ... + \theta_{t} e^{t}.
\end{equation}

@@

Comparing \ref{mdof_form_components} and \ref{mdof_Ham_eq} we see that the time components $\omega_{at}$ are actually the components of $\vec{S}$. In fact, the only non-trivial components of $\omega$ are components of $\vec{S}$. These two objects, then, are not only related: they express the same information.

For a single degree of freedom, recalling that $S^t = 1$, we can describe $\omega$ with the following nine components
\begin{equation}\label{mdof_reduceto_sdof}
	\omega_{ab}=(-\partial \wedge\theta)_{ab} = \begin{bmatrix}
		0 & S^t & -S^p \\
		-S^t & 0 & S^q \\
		S^p & - S^q & 0
	\end{bmatrix}.
\end{equation}
In three dimensions, using the Levi-Civita symbol $\epsilon^{abc}$ we can write
\begin{equation}
	S^a = \epsilon^{abc} \omega_{bc}
\end{equation}
which tells us that $\vec{S}$ is the pseudo-vector corresponding to the direction perpendicular to $\omega$. This is the same relationship that we would have, for example, for the magnetic field psuedo-vector and the spatial components of the electromagnetic field in relativity. If we express $\omega$ in terms of $\vec{\theta}$, we have
\begin{equation}
	S^a = - \epsilon^{abc} \partial_b \theta_c
\end{equation}
which recovers \ref{sdof_displacement_potential}: the displacement field is minus the curl of $\omega$. Thus we can see exactly how the general case reduces to the single d.o.f. case.

For (iv), we simply need to apply Stokes' theorem. The situation is identical to the single degree of freedom, except that the flow now is not given by the surface integral of $\vec{S}$, but of the form $\omega$. The path $\gamma$ and its variation $\gamma'$ will form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\omega$ is closed, its surface integral depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Thus we have
\begin{equation}
\tag*{(\ref{mdof_action})}
\begin{aligned}
	- \int_{\Sigma} \omega(d\Sigma) &=
	\int_{\Sigma} \partial \wedge \theta (d\Sigma) \\
	&= \oint_{\partial \Sigma = \gamma \cup \gamma'} \theta(d\gamma) \\
	&= \int_{\gamma} \theta(d\gamma) - \int_{\gamma'} \theta (d\gamma') \\
	&= \delta \int_{\gamma} \theta (d\gamma).
\end{aligned}
\end{equation}
